---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MOTL

<!-- badges: start -->
<!-- badges: end -->

MOTL infers latent factor values for a multi-omics target dataset, consisting 
of a small number of samples, by incorporating latent factor values already 
inferred with a MOFA factorization of a large, heterogeneous, learning dataset.



## Installation

You can install the development version of MOTL from 
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
# devtools::install_github("MOohTus/MOTL-pkg")
```

## How to use MOTL

### Preprocessing of learning dataset factorization

```{r learningPrep}
LrnDir <- "LrnData"
```

### Preproessing of target dataset factorization

```{r targetPrep}
YTrg_list <- list(
  mRNA = "expdat_mRNA",
  miRNA = "expdat_miRNA",
  DNAme = "expdat_DNAme",
  SNV = "expdat_SNV"
)
```

### Transfer learning factorization with MOTL

```{r MOTL}
TL_data <- ""
```
