---
title: "How to run MOTL: basic commands"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to run MOTL: basic commands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  echo = TRUE
)
```

> ⚠️ Don't use the data inside the MOTL package to perform analysis.

You will find here an example that illustrate **how to use MOTL** with the basics commands. Data used in this example are subsets of the original data. Original data can be found [ZENODO]<ZENODO>. 

Main steps to perform transfer learning analysis using MOTL: 

- Prepare learning dataset ``Lrn``
- Prepare target dataset ``Trg``
- Define parameters for transfer learning ``TL_param``
- Run transfer learning 

## Load libraries

First, load the necessary packages: 

- `MOTL` for the transfer learning analysis
- `MOFA2` to read the learning factorized model

```{r libraries, message = FALSE, warning = FALSE}
library("MOTL")
library("MOFA2")
```

## Learning dataset factorization

Renvoyer à comment on fait le learning dataset factorization ? 
Import the learning dataset and the corresponding factorization model:

```{r}
data("Lrn", package = "MOTL")
```


### Learning dataset metadata

Learning dataset metadata are stored in ``Lrn$Lrn_meta`` object in the MOTL package. 
```{r Lrn_loadData}
expdat_meta_Lrn <- Lrn$Lrn_meta
```

You can also load them from .rds file:
```{r Lrn_metadata-1, eval = FALSE}
expdat_meta_Lrn <- readRDS("expdat_meta.rds")
```

Faire une description des informations contenues dans l'objet / The `expdat_meta_Lrn` variable contains information about the learning dataset construction. 
```{r Lrn_metadata-2}
names(expdat_meta_Lrn)
```

You can retrieve for example mRNA feature names
```{r Lrn_metadata-3}
expdat_meta_Lrn$ftrs_mRNA[c(1:5)]
```

You can retrieve for example SNV feature names
```{r Lrn_metadata-4}
expdat_meta_Lrn$ftrs_SNV[c(1:5)]
```

You can retrieve for example sample names
```{r Lrn_metadata-5}
expdat_meta_Lrn$smpls[c(1:5)]
```

### Learning dataset factorization model

The learning dataset factorization model is stored in `Lrn$Fctrzn`:
```{r}
Fctrzn <- Lrn$Fctrzn
```

You can also load the model from the output file from MOFA:
```{r, eval = FALSE}
inputModel <- file.path("path/of/model", "Model.hdf5")
Fctrzn <- MOFA2::load_model(file = inputModel)
```

```{r Lrn_fctrzn-1}
Fctrzn
```

- 4 views were used for the factorization: `mRNA`, `miRNA`, `DNAme` and `SNV`.  
- `mRNA`, `DNAme` and `SNV` datasets contain 1000 features each and `miRNA` 250 features.
- Only one group was define for the factorization
- Datasets contain 250 samples (shared between each view)
- Factorization of the learning dataset found 20 factors

See MOFA2 doc pour plus d'info sur ce que contient exactement l'objet de MOFA2
Dire que cet objet vient de MOFA2

### Initialise values from factorization

- `viewsLrn` views of the learning dataset
```{r init_viewsLrn}
viewsLrn <- Fctrzn@data_options$views
viewsLrn
```

- `likelihoodsLrn` le machin de chaque view:
```{r init_likeLrn}
likelihoodsLrn <- Fctrzn@model_options$likelihoods
likelihoodsLrn
```

- 
```{r init_MLrn}
MLrn <- Fctrzn@dimensions$M
MLrn
```

Expliquer pourquoi on fait ça:

```{r}
CenterTrg <- FALSE
```



```{r init_valuesLrn-1, eval = FALSE}
Fctrzn@expectations[["Tau"]] <- Tau_init(viewsLrn, Fctrzn, InputModel)
Fctrzn@expectations[["TauLn"]] <- sapply(viewsLrn, TauLn_calculation, likelihoodsLrn, Fctrzn, LrnFctrnDir)
Fctrzn@expectations[["WSq"]] <- sapply(viewsLrn, WSq_calculation, Fctrzn, LrnFctrnDir)
Fctrzn@expectations[["W0"]] <- sapply(viewsLrn, W0_calculation, CenterTrg, Fctrzn, LrnFctrnDir)
```

CenterTrg parler de ce parametre


Initialized data are stored in the `Lrn$Fctrzn_init` object: 
```{r init_valuesLrn}
Fctrzn <- Lrn$Fctrzn_init
```


## Target dataset factorization

```{r}
data("Trg")
```


### Read files

Target data are stored in the `Trg$YTrg_list` object:

Donner le format des fichiers

```{r}
YTrg_list <- Trg$YTrg_list
```

You can load each view of the target dataset from .csv file: 
```{r, eval = FALSE}
expdat_mRNA <- read.table("mRNA.csv", sep = ",")
expdat_mRNA <- as.matrix(expdat_mRNA)
```

For genes features, ça doit être dans ce format car le format de Lrn c'est comme ça. ON a fait une fonction pour ça : 
```{r, eval = FALSE}
expdat_mRNA = mRNA_addVersion(expdat = expdat_mRNA, Lrndat = Fctrzn@expectations$W$mRNA)
```


Target dataset must be in matrix. 

Then, each view are store in a list:
```{r, eval = FALSE}
YTrg_list <- list(
  mRNA = expdat_mRNA,
  miRNA = expdat_miRNA,
  DNAme = expdat_DNAme,
  SNV = expdat_SNV
)
```

- sample in columns
- features in rows
- samples columns should be same in same order

```{r}
YTrg_list$mRNA
```

```{r}
YTrg_list$miRNA
```

```{r}
YTrg_list$DNAme
```

```{r}
YTrg_list$SNV[c(1:5), c(1:5)]
```

```{r}
lapply(YTrg_list, dim)
```

### Prepare target dataset

Make normalisation and/or transformation
Harmonize with Lrn (have same features and same order)
Order names between views

```{r}
smpls <- colnames(YTrg_list[[1]])
viewsTrg <- names(YTrg_list)
views <- viewsLrn[is.element(viewsLrn,viewsTrg)]

```


```{r, message = FALSE}
YTrg_prep <- TCGATargetDataPreparation(views = views,
                                       YTrgFull = YTrg_list,
                                       brcds_SS = Trg$brcds_SS,
                                       SS = 1,
                                       Fctrzn = Fctrzn,
                                       smpls = smpls,
                                       normalization = FALSE,
                                       expdat_meta_Lrn = expdat_meta_Lrn,
                                       transformation = FALSE
)
```

```{r}
YTrg_prep$mRNA[c(1:5), c(1:5)]
```

```{r}
YTrg_prep$miRNA[c(1:5), c(1:5)]
```

```{r}
YTrg_prep$DNAme[c(1:5), c(1:5)]
```

```{r}
YTrg_prep$SNV[c(1:5), c(1:5)]
```

## Initialize values for transfer learning

```{r}
likelihoods <- likelihoodsLrn[views]
```

```{r, message = FALSE}
TL_param <- initTransferLearningParamaters(YTrg = YTrg_prep, views = views, 
                                          Fctrzn = Fctrzn, likelihoods = likelihoods)
```

Dire de quoi il est composé.


## TRANSFERT LEARNING WITH MOTL

```{r}
ss_start_time <- Sys.time()
minFactors <- 13 ## floor when dropping factors - number of samples in evaluations
StartDropFactor <- 1 # after which iteration to start dropping factors
FreqDropFactor <- 1 ## how often to drop factors
StartELBO <- 1 #which iteration to start checking ELBO on, excl initiation iteration
FreqELBO <- 5 #how often to assess the ELBO
DropFactorTH <- 0.01 # factor with lowest max variance, that is less than this, is dropped
MaxIterations <- 10
MinIterations <- 2 # as per MOFA with defaults in python, >= 2 (hardcoded floor), exl initial setup
ConvergenceIts <- 2 # as per MOFA python defaults
ConvergenceTH <- 0.0005 # default for MOFA - fast option
PoisRateCstnt <- 0.0001 ## amount to add to the poison rate function to avoid errors
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
TL_data <- transferLearning_function(TL_param = TL_param, MaxIterations = MaxIterations, MinIterations = MinIterations,
                                    minFactors = minFactors, StartDropFactor = StartDropFactor, 
                                    FreqDropFactor = FreqDropFactor, StartELBO = StartELBO, 
                                    FreqELBO = FreqELBO, DropFactorTH = DropFactorTH,
                                    ConvergenceIts = ConvergenceIts, ConvergenceTH = ConvergenceTH,
                                    CenterTrg = CenterTrg, ss_start_time = ss_start_time,
                                    views = views, likelihoods = likelihoods, Fctrzn = Fctrzn)
```

```{r, message = FALSE, warning = FALSE, eval = FALSE}
TL_data <- transferLearning_function(TL_param = TL_param, MaxIterations = MaxIterations, MinIterations = MinIterations,
                                    minFactors = minFactors, StartDropFactor = StartDropFactor, 
                                    FreqDropFactor = FreqDropFactor, StartELBO = StartELBO, 
                                    FreqELBO = FreqELBO, DropFactorTH = DropFactorTH,
                                    ConvergenceIts = ConvergenceIts, ConvergenceTH = ConvergenceTH,
                                    CenterTrg = CenterTrg, ss_start_time = ss_start_time,
                                    views = views, likelihoods = likelihoods, Fctrzn = Fctrzn)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
file.remove("TL_data.rds")
```

Results are saved into .rds in the `outputDir`. 

Then have access to the results:
```{r}
ZMu = TL_data$ZMu
W_mRNA = TL_data$Fctrzn_Lrn_W$mRNA
```

```{r}
dim(W_mRNA)
W_mRNA[c(1:5), c(1:5)]
```


```{r}
sessionInfo()
```
