---
title: "How to run MOTL: basic example"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{How to run MOTL: basic example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  echo = TRUE
)
```

# Introduction 

This is an example of **how to use MOTL** with the basics commands. Data used in this example are subsets of the original data. 

The complete learning dataset can be found via this link: [zenodo link](https://zenodo.org/records/10848217).

## Toy datasets in MOTL

Datasets used in this example are stored in two objects:

- [Lrn](../doc/reference/Lrn.html): learning dataset (used for transfer learning)
- [Trg](../doc/reference/Trg.html): target dataset (data to analyse)

⚠️ Don't use the data inside the MOTL package to perform analysis.

## Main steps

Main steps to perform transfer learning analysis using MOTL:

1. Initialize the learning dataset ``Lrn``
2. Prepare the target dataset ``Trg``
3. Define parameters for transfer learning ``TL_param``
4. Run [`transferLearning_function()`](../docs/reference/transferLearning_function.html)

## Load libraries

First, load the necessary packages: 

- `MOTL` for the transfer learning analysis
- `MOFA2` to read the learning factorized model

```{r libraries, message = FALSE, warning = FALSE}
library("MOTL")
library("MOFA2")
```

## Learning dataset factorization

Load the learning dataset and the corresponding factorization model from the `MOTL` package.
```{r}
data("Lrn", package = "MOTL")
```

### Learning dataset metadata

#### Load from `Lrn`

The learning dataset metadata are stored in ``Lrn$Lrn_meta``. 
```{r Lrn_loadData}
expdat_meta_Lrn <- Lrn$Lrn_meta
```

#### Load from a file

You can also load them from .rds file:
```{r Lrn_metadata-1, eval = FALSE}
expdat_meta_Lrn <- readRDS("expdat_meta.rds")
```

#### `expdat_meta_Lrn` object

The `expdat_meta_Lrn` object contains information about the learning dataset construction.
```{r Lrn_metadata-2}
names(expdat_meta_Lrn)
```

You can retrieve for example mRNA feature names.
```{r Lrn_metadata-3}
expdat_meta_Lrn$ftrs_mRNA[c(1:5)]
```

You can also retrieve the SNV feature names.
```{r Lrn_metadata-4}
expdat_meta_Lrn$ftrs_SNV[c(1:5)]
```

Or, you can retrieve the sample names.
```{r Lrn_metadata-5}
expdat_meta_Lrn$smpls[c(1:3)]
```

### Learning dataset factorization model

#### Load from `Lrn`

The learning dataset factorization model is stored in `Lrn$Fctrzn`.
```{r Lrn_model-1}
Fctrzn <- Lrn$Fctrzn
```

#### Load from a file

You can also load the model from the output file from MOFA.
```{r Lrn_model-2, eval = FALSE}
inputModel <- file.path("path/of/model", "Model.hdf5")
Fctrzn <- MOFA2::load_model(file = inputModel)
```

#### `Fctrzn` object

The `Fctrzn` was created using `MOFA2` package and is composed of: 

- 4 views, uses for the factorization: `mRNA`, `miRNA`, `DNAme` and `SNV`,
- `mRNA`, `DNAme` and `SNV` datasets with 1000 features each and `miRNA` with 250 features,
- only one group, define for the factorization,
- datasets with 250 samples (shared between each view),
- 20 factors, found during the factorization of the learning dataset.

```{r Lrn_fctrzn-1}
Fctrzn
```

See `?MOFA2::MOFA` for more details about the `MOFA` object.

### Initialise values from factorization

- `viewsLrn`: views of the learning dataset
```{r init_viewsLrn}
viewsLrn <- Fctrzn@data_options$views
viewsLrn
```

- `likelihoodsLrn`: defined likelihoods of each view
```{r init_likeLrn}
likelihoodsLrn <- Fctrzn@model_options$likelihoods
likelihoodsLrn
```

- `MLrn`: dimension (number of views) of the learning dataset
```{r init_MLrn}
MLrn <- Fctrzn@dimensions$M
MLrn
```

- `CenterTrg` allows the user to center the target dataset during preprocessing or leave it uncentered and use the estimated learning dataset intercepts (for normalization). Here, we will use the estimated learning dataset intercepts.
```{r}
CenterTrg <- FALSE
```

- Init the factorization expectations values
```{r init_valuesLrn-1, eval = FALSE}
Fctrzn@expectations[["Tau"]] <- Tau_init(viewsLrn, Fctrzn, InputModel)
Fctrzn@expectations[["TauLn"]] <- sapply(viewsLrn, TauLn_calculation, likelihoodsLrn, Fctrzn, LrnFctrnDir)
Fctrzn@expectations[["WSq"]] <- sapply(viewsLrn, WSq_calculation, Fctrzn, LrnFctrnDir)
Fctrzn@expectations[["W0"]] <- sapply(viewsLrn, W0_calculation, CenterTrg, Fctrzn, LrnFctrnDir)
```

Initialized data are stored in the `Lrn$Fctrzn_init` object. The following line replaces previous 4 lines. 
```{r init_valuesLrn}
Fctrzn <- Lrn$Fctrzn_init
```

## Target dataset factorization

Load the target dataset and the corresponding metadata from the `MOTL` package.
```{r}
data("Trg", package = "MOTL")
```

### Target dataset 

#### Load from `Trg`

Target dataset are stored in the `Trg$YTrg_list` object.
```{r Trg_loadData-1}
YTrg_list <- Trg$YTrg_list
```

#### Load from a file

You can load each view of the target dataset from .csv file: 
- Samples are in columns, 
- Features are in rows,
- Samples columns should be in same order,
- Target dataset must be in matrix.
```{r Trg_loadData-2, eval = FALSE}
expdat_mRNA <- read.table("mRNA.csv", sep = ",")
expdat_mRNA <- as.matrix(expdat_mRNA)
```

The learning dataset contains ENSEMBL ID for counts data. The ENSEMBL ID contain version of annotation (e.g. `ENSG00000122133.17`). Feature names need to be similar between target dataset and learning dataset. It possible to add the annotation version from the learning dataset to the target dataset to have consistent names using the `mRNA_addVersion()` function. 
```{r mRNAversion, eval = FALSE}
expdat_mRNA = mRNA_addVersion(expdat = expdat_mRNA, Lrndat = Fctrzn@expectations$W$mRNA)
```

Then, each view is stored in a list:
```{r Trg_loadData-3, eval = FALSE}
YTrg_list <- list(
  mRNA = expdat_mRNA,
  miRNA = expdat_miRNA,
  DNAme = expdat_DNAme,
  SNV = expdat_SNV
)
```

##### mRNA data
Because the toy data was created using the TCGA data, mRNA is in `SummarizedExperiment` object instead of a `matrix`.
```{r Trg_list-mRNA}
YTrg_list$mRNA
```

##### miRNA data
Because the toy data was created using the TCGA data, miRNA is in `SummarizedExperiment` object instead of a `matrix`.
```{r Trg_list-miRNA}
YTrg_list$miRNA
```

##### DNAme data
Because the toy data was created using the TCGA data, DNAme is in `SummarizedExperiment` object instead of a `matrix`.
```{r Trg_list-DNAme}
YTrg_list$DNAme
```

##### SNV data
```{r Trg_list-SNV}
YTrg_list$SNV[c(1:5), c(1:3)]
```

##### YTrg list 
```{r Trg_list}
lapply(YTrg_list, dim)
```

- mRNA: ```r nrow(YTrg_list$mRNA)``` features and ```r ncol(YTrg_list$mRNA)``` samples
- miRNA: ```r nrow(YTrg_list$miRNA)``` features and ```r ncol(YTrg_list$miRNA)``` samples
- DNAme: ```r nrow(YTrg_list$DNAme)``` features and ```r ncol(YTrg_list$DNAme)``` samples
- SNV: ```r nrow(YTrg_list$SNV)``` features and ```r ncol(YTrg_list$SNV)``` samples

### Prepare target dataset

Extract sample and view names in variables. 
```{r Trg_preprocessing-1}
smpls <- colnames(YTrg_list[[1]])
viewsTrg <- names(YTrg_list)
views <- viewsLrn[is.element(viewsLrn,viewsTrg)]
```

To prepare the target dataset, use the `TCGATargetDataPreparation()`: 
- normalize and/or transform
- harmonize with Lrn (have same features and same order)
- order names between views

Here, we will no transform (`transformation = FALSE`) neither normalize dataset (`normalization = FALSE`).
```{r Trg_preprocessing-2, message = FALSE}
YTrg_prep <- TCGATargetDataPreparation(views = views,
                                       YTrgFull = YTrg_list,
                                       brcds_SS = Trg$brcds_SS,
                                       SS = 1,
                                       Fctrzn = Fctrzn,
                                       smpls = smpls,
                                       normalization = FALSE,
                                       expdat_meta_Lrn = expdat_meta_Lrn,
                                       transformation = FALSE
)
```

#### mRNA after preparation
```{r}
YTrg_prep$mRNA[c(1:5), c(1:3)]
```

#### miRNA after preparation
```{r}
YTrg_prep$miRNA[c(1:5), c(1:3)]
```

#### DNAme after preparation
```{r}
YTrg_prep$DNAme[c(1:5), c(1:3)]
```

#### SNV after preparation
```{r}
YTrg_prep$SNV[c(1:5), c(1:3)]
```

## Initialize values for transfer learning

Retrieve view likelihoods of the learning dataset.
```{r init-TL-1}
likelihoods <- likelihoodsLrn[views]
```

Prepare the input of the transfer learning:
- `YTrg`: list of the target dataset matrices (prepared with `TCGATargetDataPreparation()`)
- `views`: vector of target dataset view names
- `Fctrzn`: the learning dataset factorization model
- `likelihoods`: list of view likelihoods

```{r init-TL-2, message = FALSE}
TL_param <- initTransferLearningParamaters(YTrg = YTrg_prep, views = views, 
                                          Fctrzn = Fctrzn, likelihoods = likelihoods)
```

```{r init-TL-3}
names(TL_param)
```

## Transfer learning with `MOTL`

```{r MOTL-input}
ss_start_time <- Sys.time()
minFactors <- 13 ## floor when dropping factors - number of samples in evaluations
StartDropFactor <- 1 # after which iteration to start dropping factors
FreqDropFactor <- 1 ## how often to drop factors
StartELBO <- 1 #which iteration to start checking ELBO on, excl initiation iteration
FreqELBO <- 5 #how often to assess the ELBO
DropFactorTH <- 0.01 # factor with lowest max variance, that is less than this, is dropped
MaxIterations <- 10
MinIterations <- 2 # as per MOFA with defaults in python, >= 2 (hardcoded floor), exl initial setup
ConvergenceIts <- 2 # as per MOFA python defaults
ConvergenceTH <- 0.0005 # default for MOFA - fast option
PoisRateCstnt <- 0.0001 ## amount to add to the poison rate function to avoid errors
```

```{r MOTL, message = FALSE, warning = FALSE, echo = FALSE}
TL_data <- transferLearning_function(TL_param = TL_param, MaxIterations = MaxIterations, MinIterations = MinIterations,
                                    minFactors = minFactors, StartDropFactor = StartDropFactor, 
                                    FreqDropFactor = FreqDropFactor, StartELBO = StartELBO, 
                                    FreqELBO = FreqELBO, DropFactorTH = DropFactorTH,
                                    ConvergenceIts = ConvergenceIts, ConvergenceTH = ConvergenceTH,
                                    CenterTrg = CenterTrg, ss_start_time = ss_start_time,
                                    views = views, likelihoods = likelihoods, Fctrzn = Fctrzn)
```

```{r MOTL-results, message = FALSE, warning = FALSE, eval = FALSE}
TL_data <- transferLearning_function(TL_param = TL_param, MaxIterations = MaxIterations, MinIterations = MinIterations,
                                    minFactors = minFactors, StartDropFactor = StartDropFactor, 
                                    FreqDropFactor = FreqDropFactor, StartELBO = StartELBO, 
                                    FreqELBO = FreqELBO, DropFactorTH = DropFactorTH,
                                    ConvergenceIts = ConvergenceIts, ConvergenceTH = ConvergenceTH,
                                    CenterTrg = CenterTrg, ss_start_time = ss_start_time,
                                    views = views, likelihoods = likelihoods, Fctrzn = Fctrzn)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
file.remove("TL_data.rds")
```

Results are saved into .rds in the `outputDir`. 

Then have access to the results:
```{r}
ZMu = TL_data$ZMu
W_mRNA = TL_data$Fctrzn_Lrn_W$mRNA
```

```{r}
dim(W_mRNA)
W_mRNA[c(1:5), c(1:5)]
```


# Session info

```{r sessionInfo}
sessionInfo()
```

# References
