---
title: "UseCase_Glioblastoma"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{UseCase_Glioblastoma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r toRun, echo = FALSE}
setwd("/home/morgane/Documents/01_Projects/03_OtherProjects/05_David_Thesis/test_package_aout25/MOTL/vignettes/")

Seed <- 1234567
set.seed(Seed)

library("MOTL")
library("MOFA2")

metadata_file <- "Glioblastoma/02_RNAseq/phenodata_RNAseq.tsv"
mRNA_dir <- "Glioblastoma/02_RNAseq/CountData/"
DNAme_dir <- "Glioblastoma/03_DNAmethylation/"

smpls_df <- read.table("Glioblastoma_metadata.txt", head = TRUE)
smpls <- smpls_df$Sample

expdat_mRNA <- as.matrix(read.table("Glioblastoma/02_RNAseq/CountData/mRNA.csv", head = TRUE, sep = ",", check.names = FALSE))
expdat_DNAme <- as.matrix(read.table("Glioblastoma/03_DNAmethylation/DNAme.csv", head = TRUE, sep = ",", check.names = FALSE))

Lrn_dir <- "Lrn_5000D/"
LrnFctrn_dir <- "Lrn_5000D_Fctrzn_100K_001TH/"

expdat_meta_Lrn <- readRDS(file.path(Lrn_dir, "expdat_meta_Lrn.rds"))
InputModel <- file.path(LrnFctrn_dir, "Model.hdf5")
Fctrzn <- load_model(file = InputModel)
CenterTrg <- FALSE
viewsLrn <- Fctrzn@data_options$views
likelihoodsLrn <- Fctrzn@model_options$likelihoods
MLrn <- Fctrzn@dimensions$M
Fctrzn@expectations[["Tau"]] <- Tau_init(viewsLrn, Fctrzn, InputModel)
Fctrzn@expectations[["TauLn"]] <- sapply(viewsLrn, TauLn_calculation, likelihoodsLrn, Fctrzn)
Fctrzn@expectations[["WSq"]] <- sapply(viewsLrn, WSq_calculation, Fctrzn)
Fctrzn@expectations[["W0"]] <- sapply(viewsLrn, W0_calculation, CenterTrg, Fctrzn, LrnFctrn_dir)

TL_outDir <- "Glioblastoma/MOTL_results"

```

Libraries used for input data preparation and transfer learning.
```{r libraries-1, eval = TRUE, message = FALSE, warning = FALSE}
library("MOTL")
library("MOFA2")
```

Helper functions created for this specific use case.
```{r functions, eval = TRUE}
mRNA_raw_import <- function(smpls, InRawCountsDir){
  # import raw counts line by line for each sample
  mRNA_trg_tmp <- strsplit(
    readLines(file.path(InRawCountsDir, paste0(smpls, ".genes.results")), n = -1), 
    split = "\t"
  )
  
  correct_length <- 7
  mRNA_trg_tmp_chk <- lapply(mRNA_trg_tmp, function(x) length(x) == correct_length)
  mRNA_trg_tmp <- mRNA_trg_tmp[unlist(mRNA_trg_tmp_chk)]
  mRNA_trg_tmp <- do.call(rbind, mRNA_trg_tmp)
  colnames(mRNA_trg_tmp) <- mRNA_trg_tmp[1,]
  mRNA_trg_tmp <- mRNA_trg_tmp[-1,]
  
  mRNA_trg_tmp <- data.frame(
    gene_id = mRNA_trg_tmp[, "gene_id"],
    sample = smpls,
    expected_count = round(as.numeric(mRNA_trg_tmp[, "expected_count"]))
  )
  
  return(mRNA_trg_tmp)
}

mRNA_stripVersion <- function(expdat){
  # strip version and gene symbol from ensembl ids
  vStart <- regexpr("[.]", rownames(expdat))
  suffixStart <- regexpr("_PAR_Y", rownames(expdat))
  suffixEnd <- suffixStart -1 + attr(regexpr("_PAR_Y", rownames(expdat)), "match.length")
  rownames(expdat) <- paste0(
    substr(rownames(expdat), 1, vStart-1),
    substr(rownames(expdat), suffixStart, suffixEnd)
  )
  # return tidied up matrix
  return(expdat)
}

DifFactTester <- function(x, minSamples, sampleGroup) {
  if (minSamples >= 4) {
    wilcox.test(x ~ sampleGroup)$p.value
  } else {
    t.test(x ~ sampleGroup)$p.value
  }
}

ScaleW <- function(x) {
  return(x/norm(matrix(x), type = "F"))
}

sigGeneSetFct <- function(v, WMat, groups){
    ## Retrieve element from the inputs
    k <- v[1]
    category <- v[2]
    subcategory <- v[3]
    rankings <- WMat[,k]
    # names(rankings) <- rownames(WMat)
    
    ## Retrieve gene sets data frame
    gene_sets <- msigdbr(species = "Homo sapiens",
                         category = category,
                         subcategory = subcategory)
    ## Reformat the gene sets data frame
    gene_sets <- gene_sets %>% 
        dplyr::distinct(gs_name, ensembl_gene) %>%
        as.data.frame()
    gene_sets <- split(x = gene_sets$ensembl_gene, f = gene_sets$gs_name)
    
    ## Run the enrichment analysis
    fgseaRes <- fgsea(pathways = gene_sets,
                      stats = rankings,
                      minSize = 15,
                      maxSize = 500)
    ## Reshape results
    fgseaRes <- fgseaRes %>%
        dplyr::select(pathway, padj, ES, NES, size) %>%
        as.data.frame()
    ## Select significant results
    fgseaRes <- fgseaRes[fgseaRes$padj < 0.01 & !is.na(fgseaRes$padj),]
    fgseaRes$k <- rep(k, nrow(fgseaRes))
    fgseaRes$subcategory <- rep(subcategory, nrow(fgseaRes))
    fgseaRes$groups <- rep(groups, nrow(fgseaRes))
    
    return(fgseaRes)
}
```

Example of real data analysis. You can find the entire analysis that you can retreive in the paper in the github paper. 

```{r setup, eval = TRUE}
Seed <- 1234567
set.seed(Seed)
```

# Import Glioblastoma dataset

Download the data from [zenodo](https://zenodo.org/records/7380252):

- mRNA data: https://zenodo.org/records/7380252/files/02_RNAseq.zip?download=1
- DNAme data: https://zenodo.org/records/7380252/files/03_DNAmethylation.zip?download=1

In this use case, data are stored in the `Glioblastoma/` folder.

```{r G_files, eval = TRUE}
metadata_file <- "Glioblastoma/02_RNAseq/phenodata_RNAseq.tsv"
mRNA_dir <- "Glioblastoma/02_RNAseq/CountData/"
DNAme_dir <- "Glioblastoma/03_DNAmethylation/"
```

## Import Glioblastoma metadata

First, import the metadata associated to datasets. 
```{r metadata-1, eval = TRUE}
smpls_df <- read.table(metadata_file, header = TRUE)
smpls_df$Sample
```

Replace `.` with `-` and remove the `X` in the sample names.
```{r metadata-2, eval = TRUE}
smpls_df$Sample <- sub("[.]", "-", smpls_df$Sample)
smpls_df$Sample <- sub("X", "", smpls_df$Sample)
smpls_df$Sample
```

We are interested in the cancer subtypes: 

- `CL`: classical
- `MS`: mesenchymal
- `PN`: proneural

Samples are ordered randomly.
```{r metadata-3, eval = TRUE}
HGCCgroups <- c("Normal", "CL", "MS", "PN")
smpls <- smpls_df$Sample[is.element(smpls_df$HGCC, HGCCgroups)]
smpls <- sample(smpls, length(smpls), replace = FALSE)
```

We selected ```r length(smpls)``` samples for the analysis.
```{r metadata-4, eval = TRUE}
smpls
```

Retrieve metadata for selected samples. Create a column with `Case` and `Normal` conditions. 
```{r metadata-5, eval = TRUE}
smpls_df <- smpls_df[match(smpls, smpls_df$Sample),]
smpls_df$CaseControl <- smpls_df$HGCC
smpls_df$CaseControl[smpls_df$CaseControl != "Normal"] <- "Case"
```

```{r metadata-6, eval = TRUE}
write.table(x = smpls_df, file = "Glioblastoma_metadata.txt", sep = "\t", 
            quote = FALSE, row.names = FALSE)
```

## Import Glioblastoma mRNA data

Import mRNA data and make a raw counts matrix.
```{r mRNA-1, eval = TRUE}
expdat_mRNA <- lapply(X = smpls, FUN = mRNA_raw_import, mRNA_dir)
expdat_mRNA <- do.call(rbind, expdat_mRNA)
expdat_mRNA <- reshape2::dcast(expdat_mRNA, formula = gene_id ~ sample)
rownames(expdat_mRNA) <- expdat_mRNA$gene_id
expdat_mRNA <- as.matrix(expdat_mRNA[, -1])
expdat_mRNA[c(1:5), c(1:7)]
```

Tidy up feature names and order sample columns.
```{r mRNA-2, eval = TRUE}
expdat_mRNA <- mRNA_stripVersion(expdat_mRNA)
expdat_mRNA <- expdat_mRNA[,smpls]
expdat_mRNA[c(1:5), c(1:7)]
```

The raw counts matrix is saved into a .csv file to facilitate the loading.
```{r mRNA-3}
write.table(x = expdat_mRNA, file = file.path(mRNA_dir, "mRNA.csv"), 
            sep = ",", row.names = TRUE, col.names = TRUE)
```

The mRNA dataset contains ``r nrow(expdat_mRNA)`` features (genes) and ``r ncol(expdat_mRNA)`` samples.

## Import Glioblastoma DNAme data

Read DNAme data file. You will get warnings because the file contains some corrupted lines. We will remove these lines after.
```{r DNAme-1, eval = TRUE, warning = FALSE}
expdat_DNAme <- strsplit(
    readLines(file.path(DNAme_dir, "5mC_bVals_HGCC.txt"), n = -1),
    split = "\t"
)
```

Reshape the DNAme data object and keep only rows with the correct number of elements.
```{r DNAme-2, eval = TRUE}
DNAme_smpls <- expdat_DNAme[[1]]
correct_length <- 15
expdat_DNAme_chk <- lapply(expdat_DNAme, function(x) length(x) == correct_length)
expdat_DNAme <- expdat_DNAme[unlist(expdat_DNAme_chk)]
expdat_DNAme <- do.call(rbind, expdat_DNAme)
```

Make sure only rows with probe IDs are kept
```{r DNAme-3, eval = TRUE}
expdat_DNAme <- 
    expdat_DNAme[is.element(substr(expdat_DNAme[,1], 1, 2), c("cg", "ch")),]
```

Move sample IDs to the row name columns and add sample names.
```{r DNAme-4, eval = TRUE}
rownames(expdat_DNAme) <- expdat_DNAme[,1]
expdat_DNAme <- expdat_DNAme[,-1]
colnames(expdat_DNAme) <- DNAme_smpls
expdat_DNAme[c(1:5), c(1:3)]
```

Then, values need to be convert into numeric values. Don't worry, the introduced NA are in the `e12` column and we will remove it.
```{r DNAme-5, eval = TRUE, warning = FALSE, message = FALSE}
expdat_DNAme <- matrix(as.numeric(expdat_DNAme), 
                       ncol = ncol(expdat_DNAme),
                       dimnames = list(rownames(expdat_DNAme),
                                       colnames(expdat_DNAme)))
```

To analysis the DNAme data, values need to be converted into M-Values.
```{r DNAme-6, eval = TRUE}
expdat_DNAme <- sesame::BetaValueToMValue(expdat_DNAme)
```

Finally, we remove samples that are not in the `smpls` variable and order sample columns. 
```{r DNAme-7, eval = TRUE}
expdat_DNAme <- expdat_DNAme[,smpls]
```

The M-Values matrix is saved into a .csv file to facilitate the loading.
```{r DNAme-8}
write.table(x = expdat_DNAme, file = file.path(DNAme_dir, "DNAme.csv"), 
            sep = ",", row.names = TRUE, col.names = TRUE)
```

The DNAme dataset contains ``r nrow(expdat_DNAme)`` features (probes) and ``r ncol(expdat_DNAme)`` samples.


```{r}
smpls_raw <- sub("-", ".", smpls)
```

# Learning dataset

Download the learning dataset factorization files from [zenodo](https://zenodo.org/records/10848217):

- TCGA dataset factorization: [Lrn_5000D_Fctrzn_100k_001TH.zip](https://zenodo.org/records/10848217/files/Lrn_5000D_Fctrzn_100K_001TH.zip?download=1)
- Metadata file from TCGA data processing: [expdat_meta.rds](https://zenodo.org/records/10848217/files/expdat_meta.rds?download=1)

Here, factorization data are stored in the `Lrn_5000D_Fctrzn_100K_001TH/` folder and metadata in the `Lrn_5000D/` folder.
```{r Lrn_files, eval = TRUE}
Lrn_dir <- "Lrn_5000D/"
LrnFctrn_dir <- "Lrn_5000D_Fctrzn_100K_001TH/"
```

## Import learning dataset metadata

Import metadata of the learning dataset creation. 
```{r Lrn-1, eval = TRUE}
expdat_meta_Lrn <- readRDS(file.path(Lrn_dir, "expdat_meta_Lrn.rds"))
```

## Import learning dataset factorization model

Import the learning dataset factorization model, created using `MOFA2`.
```{r Lrn-2, eval = TRUE}
InputModel <- file.path(LrnFctrn_dir, "Model.hdf5")
Fctrzn <- load_model(file = InputModel)
```

## Prepare learning dataset for TL

For the preprocessing step of the target dataset, you have two choices:

- `CenterTrg <- TRUE`: center the target dataset values
- `CenterTrg <- FALSE`: leave the values uncentered and use instead the estimated learning dataset intercepts

```{r Lrn-3, eval = TRUE}
CenterTrg <- FALSE
```

Initialize values from learning dataset factorization:

- `viewsLrn`: names of the learning dataset view (e.g. `r Fctrzn@data_options$views`)
- `likelihoodsLrn`: the defined likelihood of each view (e.g. `r Fctrzn@model_options$likelihoods`)
- `MLrn`: number of views in the learning dataset (`r Fctrzn@dimensions$M`)

```{r Lrn-4, eval = TRUE}
viewsLrn <- Fctrzn@data_options$views
likelihoodsLrn <- Fctrzn@model_options$likelihoods
MLrn <- Fctrzn@dimensions$M
```

In this use case, to initialize or caluclate the expected values, we use values from the leaning dataset factorization model generated using MOFA2.

- Initialize the `Tau` values using the one from the learning dataset model
- The log(Tau) values `TauLn` are calculated using the `Tau` values
- The squared weight values `WSq` are calculated using the weight values `W` 
- The weight intercept values `W0` are loaded from the estimated intercept file (`CenterTrg <- FALSE`) 

```{r Lrn-5, eval = TRUE}
Fctrzn@expectations[["Tau"]] <- Tau_init(viewsLrn, Fctrzn, InputModel)
Fctrzn@expectations[["TauLn"]] <- sapply(viewsLrn, TauLn_calculation, likelihoodsLrn, Fctrzn)
Fctrzn@expectations[["WSq"]] <- sapply(viewsLrn, WSq_calculation, Fctrzn)
Fctrzn@expectations[["W0"]] <- sapply(viewsLrn, W0_calculation, CenterTrg, Fctrzn, LrnFctrn_dir)
```

# Target dataset

First, feature names need to be the same as the one that are in the learning dataset. We will add gene version names that are in the learning dataset to the mRNA dataset. Here, some of the features are not retrieve in the learning dataset, so an `NA` is added instead of a number. These features will be removed after and not used in the analysis.

```{r init_TL-1, eval = TRUE}
expdat_mRNA <- mRNA_addVersion(expdat = expdat_mRNA, Lrndat = Fctrzn@expectations$W$mRNA)
expdat_mRNA[c(1:5), c(1:7)]
```

Then, we create a list of the omics data we want to analysis together: `mRNA` and `DNAme` data.
```{r init_TL-2, eval = TRUE}
YTrg_list <- list(mRNA = expdat_mRNA, DNAme = expdat_DNAme)
```

The learning and target datasets need to contain the same set of omics for transfer learning. So we define the `views` and `likelihoods` variables that contain the commun views between the two datasets and the corresponding likelihoods. 
```{r init_TL-3, eval = TRUE}
viewsTrg <- names(YTrg_list)
views <- viewsLrn[is.element(viewsLrn, viewsTrg)]
likelihoods <- likelihoodsLrn[views]
```

Target dataset are then prepared: 

- features with variance equal to zero are removed
- features are harmonized between learning and target datasets, they should contain same features
- samples are ordered in same way between views
- counts data (i.e. mRNA and miRNA) are normalized using ... (`normalization = "LrnGeoMeans"`)
- counts data (i.e. mRNA and miRNA) are log2 transformed (`transformation = TRUE`)

```{r init_TL-4, eval = TRUE}
YTrg_list <- TargetDataPreparation(views = views, YTrg_list = YTrg_list, 
                                  Fctrzn = Fctrzn, smpls = smpls, expdat_meta_Lrn = expdat_meta_Lrn,
                                  normalization = "LrnGeoMeans", transformation = TRUE)
```

Finally, we create the `TL_param` that contains all the necessary input for the transfer learning analysis

- `YTrg`: named list of the omics
- `views`: vector of view names, shared between learning and target dataset
- `Fctrzn`: learning dataset factorization model initialized
- `likelihoods`: vector of view likelihoods
```{r init_TL-5, eval = TRUE}
TL_param <- initTransferLearningParamaters(YTrg = YTrg_list, 
                                           views = views,
                                           Fctrzn = Fctrzn, 
                                           likelihoods = likelihoods
                                          )
names(TL_param)
```

The `TL_param` contain:

- `YTrg`: named list of prepared `mRNA` and `DNAme` data matrices
- `Fctrzn_Lrn_W0`: named list of weight intercept values `W0` for each view
- `Fctrzn_Lrn_W`: named list of weight values `W` for each view
- `Fctrzn_Lrn_WSq`: named list of squared weight values `WSq` fir each view
- `Tau`: named list of Tau values `Tau` for each view
- `TauLn`: named list of log(Tau) `TauLn` for each view

# Transfer learning using MOTL

Mettre une phrase de contexte, avec le nombre d'échantillon et de features dans chaque views.

Create an output folder. 
```{r MOTL-1, eval = TRUE}
TL_outDir <- "Glioblastoma/MOTL_results"
if(!dir.exists(TL_outDir)){
  dir.create(TL_outDir, showWarnings = FALSE, recursive = TRUE)
}
```

Transfer learning parameters:

- `minFactors`: floor when dropping factors - number of samples in evaluations
- `StartDropFactor`: after which iteration to start dropping factors
- `FreqDropFactor`: how often to drop factors
- `StartELBO`: which iteration to start checking ELBO on, exclude initiation iteration
- `FreqELBO`: how often to assess the ELBO
- `DropFactorTH`: factor with lowest max variance, that is less than this, is dropped
- `MaxIterations`: maximum iteration number
- `MinIterations`: minimum iteration number - at least 2 and exclude initial setup (`2` is default in MOFA)
- `ConvergenceIts`: number of consecutive iterations that change in ELBO is (`2` is default in MOFA)
- `ConvergenceTH`: threshold number for change in ELBO for checking convergence (`0.0005` is default in MOFA, correspond to the fast option)
- `PoisRateCstnt`: amount to add to the poison rate function to avoid errors(`1e-04` default)
```{r MOTL-2, eval = TRUE}
minFactors <- 13
StartDropFactor <- 1 
FreqDropFactor <- 1 
StartELBO <- 1 
FreqELBO <- 5
DropFactorTH <- 0.01 
MaxIterations <- 10000
MinIterations <- 2 
ConvergenceIts <- 2 
ConvergenceTH <- 0.0005 
```

Run MOTL. Results will be saved into the `TL_outDir` folder in the `TL_VI_data.rds` file.
```{r MOTL-3, message = FALSE}
TL_data <- transferLearning_function(TL_param = TL_param, views = views, Fctrzn = Fctrzn,
                                    likelihoods = likelihoods,
                                    MaxIterations = MaxIterations, 
                                    MinIterations = MinIterations, minFactors = minFactors, 
                                    StartDropFactor = StartDropFactor, FreqDropFactor = FreqDropFactor, 
                                    StartELBO = StartELBO, FreqELBO = FreqELBO, DropFactorTH = DropFactorTH, 
                                    ConvergenceIts = ConvergenceIts, ConvergenceTH = ConvergenceTH, 
                                    CenterTrg = CenterTrg, outputDir = TL_outDir)
```

Then, save the metadata of the transfer learning analysis.
```{r MOTL-4}
TL_meta_data <- list(
  "Seed" = Seed,
  "CenterTrg" = CenterTrg,
  "StartDropFactor" = StartDropFactor,
  "FreqDropFactor" = FreqDropFactor,
  "DropFactorTH" = DropFactorTH,
  "minFactors" = minFactors,
  "MinIterations" = MinIterations,
  "MaxIterations" = MaxIterations,
  "StartELBO" = StartELBO,
  "FreqELBO" = FreqELBO,
  "ConvergenceTH" = ConvergenceTH,
  "ConvergenceIts" = ConvergenceIts
)
saveRDS(TL_meta_data, file.path(TL_outDir, "TL_VI_meta_data.rds"))
```

# Downstream analysis

For the downstream analysis, you will need the following libraries.

```{r libraries-2, eval = TRUE, message = FALSE, warning = FALSE}
## Enrichment analysis
library("fgsea")
library("msigdbr")
## Plots
library("ggplot2")
library("pheatmap")
library("RColorBrewer")
```

## MOTL results

Set the seed to have comparable results with this current vignette.
```{r setup-2, eval = TRUE}
Seed = 1234567
mode(Seed) = 'integer'
set.seed(Seed)
```

Load the transfer learning results and corresponding metadata.
```{r down-2, eval = TRUE}
TL_outDir <- "Glioblastoma/MOTL_results"
TL_data <- readRDS(file = file.path(TL_outDir, "TL_data.rds"))
TL_meta_data <- readRDS(file = file.path(TL_outDir, "TL_VI_meta_data.rds"))
```

First, init the necessary variables.
```{r down-3, eval = TRUE}
YGauss <- TL_data$YGauss
ZMu_0 <- TL_data$ZMu_0
ZMu <- TL_data$ZMu
Fctrzn_Lrn_W0 <- TL_data$Fctrzn_Lrn_W0
Fctrzn_Lrn_W <- TL_data$Fctrzn_Lrn_W
views <- names(YGauss)
factorNames <- colnames(ZMu)
VarExpl <- TL_data$VarExpl
smpls <- rownames(TL_data$YTrgSS$mRNA)
```

Load the Glioblastoma metadata.
```{r down-1, eval = TRUE}
smpls_df <- read.table("Glioblastoma_metadata.txt", head = TRUE)
smpls_df <- smpls_df[order(smpls_df$CaseControl, smpls_df$HGCC, smpls_df$Sample),]
sampleGroup <- as.factor(smpls_df$CaseControl)
smpls_df$CaseControl[smpls_df$CaseControl == "Case"] <- "pd-GBSC"
smpls_df <- smpls_df[match(smpls, smpls_df$Sample),]
```

## Factor caracterisation: Weight matrices

## Variance explained

```{r var-1, eval = TRUE}
Z_mRNA_rel <- which(VarExpl[, "mRNA"] > 0.01)
Z_DNAme_rel <- which(VarExpl[, "DNAme"] > 0.01)
Variances_df <- reshape2::melt(TL_data$VarExpl, value.name = "Variance")
```

```{r var-2, eval = TRUE}
min_r2 <- 0.01
max_r2 <- max(Variances_df$Variance)
```

```{r var-3, eval = TRUE}
ggplot(Variances_df, aes(x = Var2, y = Var1)) + 
    geom_tile(aes_string(fill = "Variance"), color = "black") + 
    labs(x = "", y = "", title = "") +
    scale_fill_gradientn(colors = c("gray97", "darkblue"), 
            guide = "colorbar", limits = c(min_r2, max_r2)) +
    guides(fill = guide_colorbar("Var. (%)")) + 
    theme(axis.text.x = element_text(size = rel(1), 
        color = "black"), axis.text.y = element_text(size = rel(1.1), 
        color = "black"), axis.line = element_blank(), axis.ticks = element_blank(), 
        panel.background = element_blank(), strip.background = element_blank(), 
        strip.text = element_text(size = rel(1)))
```

Petit problème ici, il y a des variances négatives ... 

```{r var-4, eval = TRUE}
ggplot(Variances_df, aes(x = Var2, y = Variance)) + 
    geom_bar(stat = "identity", fill = "deepskyblue4", color = "NA", width = 0.9) +
    xlab("") + ylab("Variance explained (%)") +
    theme(axis.ticks.x = element_blank(), 
            axis.text.x = element_text(size = rel(1.1), color = "black"), 
            axis.text.y = element_text(size = rel(1), color = "black"), 
            axis.title.y = element_text(size = rel(1), color = "black"), 
            axis.line = element_line(size = rel(1), color = "black"), 
            panel.background = element_blank(), strip.background = element_blank(), 
            strip.text = element_text(size = rel(1)))
```

### Differentially active factors

Qu'est-ce que c'est ? Pourquoi on cherche ça. 

```{r diff-1, eval = TRUE}
minSamples <- min(unlist(table(smpls_df$CaseControl)))
Z_pv <- apply(X = ZMu, MARGIN = 2, FUN = DifFactTester, minSamples = minSamples, sampleGroup = smpls_df$CaseControl)
Z_pv <- p.adjust(Z_pv, method = "BH")
Z_pv_sig <- which(Z_pv < 0.05)
```

There are ``r length(Z_pv_sig)`` differential active factors. 
```{r diff-2, eval = TRUE}
Z_pv_sig
```

### Enrichment analysis

For mRNA view, we will perform an enrichment analysis of the differentially active factors using GSEA.

```{r gsea-1, eval = TRUE}
groupsTested <- paste(levels(sampleGroup), collapse = "_vs_")
W_mRNA <- t(apply(Fctrzn_Lrn_W[["mRNA"]], 1, ScaleW))
W_mRNA <- mRNA_stripVersion(W_mRNA)
```

We have ``r length(Z_pv_sig)`` differentially active factors and ``r length(Z_mRNA_rel)`` factors that explained at least 0.001 of the total variance. We select factors that are differentially active and also explained at least 0.001 of the total variance.
```{r gsea-2, eval = TRUE}
kToTest <- names(Z_pv_sig)[is.element(names(Z_pv_sig), names(Z_mRNA_rel))]
kToTest
```

We want to perform an enrichment analysis using REACTOME, KEGG, Biological Process and Cellular Component from Gene Ontology.
```{r gsea-3, eval = TRUE}
genesetCatToTest <- c("C2", "C2", "C5", "C5")
genesetSubCatToTest <- c("CP:KEGG", "CP:REACTOME", "GO:BP", "GO:CC")
```

We create the input data frame for the enrichment analysis.
```{r gsea-4, eval = TRUE}
toTest <- data.frame(
    k = rep(kToTest, each = length(genesetCatToTest)),
    category = rep(genesetCatToTest, times = length(kToTest)),
    subcategory = rep(genesetSubCatToTest, times = length(kToTest))
)
head(toTest)
```

Reshape the enrichment analysis results.
```{r gsea-5, eval = TRUE}
sigGeneSets <- apply(X = toTest, MARGIN = 1, FUN = sigGeneSetFct, WMat = W_mRNA, groups = groupsTested)
sigGeneSets <- do.call(rbind, sigGeneSets)
```

Write results as .csv
```{r gsea-6, eval = TRUE}
write.table(sigGeneSets,
            file.path(TL_outDir, paste0("SigGeneSets_", groupsTested, ".csv")),
            quote = FALSE, sep = "," , na = "", row.names = FALSE, col.names = TRUE)
```

## Z matrix analysis

Ca correspond à quoi la matrice Z

We will visualize the Z matrix using an heatmap. 

Create the input data frame of the heatmap. 
```{r z-1, eval = TRUE}
Zmu_MOTL <- t(ZMu[smpls_df$Sample,])
Zmu_MOTL[c(1:5), c(1,7)]
```

Create the corresponding annotation file.
```{r z-2, eval = TRUE}
heatMapAnno <- data.frame(SubType = as.factor(smpls_df$HGCC), Type = as.factor(smpls_df$CaseControl))
heatMapAnno$SubType <- relevel(heatMapAnno$SubType, ref = "Normal")
heatMapAnno$Type <- relevel(heatMapAnno$Type, ref = "Normal")
rownames(heatMapAnno) <- smpls_df$Sample
head(heatMapAnno)
```

Samples will be colored by the subtype of the cancer and also by the condition.
```{r z-3, eval = TRUE}
annot_colors <- list(SubType = brewer.pal(5, "Set2")[c(1,3,4,5)],
                     Type = brewer.pal(5, "Set2")[1:2])
names(annot_colors$SubType) <- levels(heatMapAnno$SubType)
names(annot_colors$Type) <- levels(heatMapAnno$Type)
```

Create the heatmap of the Z matrix.
```{r z-4, eval = TRUE}
pheatmap(Zmu_MOTL,
         border_color = NA,
         angle_col = 45,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         treeheight_row = 0,
         treeheight_col = 10,
         legend = FALSE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         annotation_col = heatMapAnno,
         annotation_colors = annot_colors,
         annotation_names_col = TRUE,
         annotation_legend = TRUE)
```

Mettre une petite explication de comment lire ça.
