---
title: "UseCase_Glioblastoma"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{UseCase_Glioblastoma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup, eval = TRUE}
Seed <- 1234567
set.seed(Seed)
```

Libraries used for input data preparation and transfer learning.
```{r libraries-1, , eval = TRUE, message = FALSE, warning = FALSE}
library("MOTL")
library("MOFA2")
```

Helper functions created for this specific use case.
```{r functions, eval = TRUE}
mRNA_raw_import <- function(smpls, InRawCountsDir){
  # import raw counts line by line for each sample
  mRNA_trg_tmp <- strsplit(
    readLines(file.path(InRawCountsDir, paste0(smpls, ".genes.results")), n = -1), 
    split = "\t"
  )
  
  correct_length <- 7
  mRNA_trg_tmp_chk <- lapply(mRNA_trg_tmp, function(x) length(x) == correct_length)
  mRNA_trg_tmp <- mRNA_trg_tmp[unlist(mRNA_trg_tmp_chk)]
  mRNA_trg_tmp <- do.call(rbind, mRNA_trg_tmp)
  colnames(mRNA_trg_tmp) <- mRNA_trg_tmp[1,]
  mRNA_trg_tmp <- mRNA_trg_tmp[-1,]
  
  mRNA_trg_tmp <- data.frame(
    gene_id = mRNA_trg_tmp[, "gene_id"],
    sample = smpls,
    expected_count = round(as.numeric(mRNA_trg_tmp[, "expected_count"]))
  )
  
  return(mRNA_trg_tmp)
}

mRNA_stripVersion <- function(expdat){
  # strip version and gene symbol from ensembl ids
  vStart <- regexpr("[.]", rownames(expdat))
  suffixStart <- regexpr("_PAR_Y", rownames(expdat))
  suffixEnd <- suffixStart -1 + attr(regexpr("_PAR_Y", rownames(expdat)), "match.length")
  rownames(expdat) <- paste0(
    substr(rownames(expdat), 1, vStart-1),
    substr(rownames(expdat), suffixStart, suffixEnd)
  )
  # return tidied up matrix
  return(expdat)
}
```

Example of real data analysis. You can find the entire analysis that you can retreive in the paper in the github paper. 

# Import Glioblastoma dataset

Download the data from [zenodo](https://zenodo.org/records/7380252):

- mRNA data: https://zenodo.org/records/7380252/files/02_RNAseq.zip?download=1
- DNAme data: https://zenodo.org/records/7380252/files/03_DNAmethylation.zip?download=1

In this use case, data are stored in the `Glioblastoma/` folder.

```{r G_files, eval = TRUE}
metadata_file <- "Glioblastoma/02_RNAseq/phenodata_RNAseq.tsv"
mRNA_dir <- "Glioblastoma/02_RNAseq/CountData/"
DNAme_dir <- "Glioblastoma/03_DNAmethylation/"
```

## Import Glioblastoma metadata

First, import the metadata associated to datasets. 
```{r metadata-1, eval = TRUE}
smpls_df <- read.table(metadata_file, header = TRUE)
smpls_df$Sample
```

Replace `.` with `-` and remove the `X` in the sample names.
```{r metadata-2, eval = TRUE}
smpls_df$Sample <- sub("[.]", "-", smpls_df$Sample)
smpls_df$Sample <- sub("X", "", smpls_df$Sample)
smpls_df$Sample
```

We are interested in the cancer subtypes: 

- `CL`: classical
- `MS`: mesenchymal
- `PN`: proneural

Samples are ordered randomly.
```{r metadata-3, eval = TRUE}
HGCCgroups <- c("Normal", "CL", "MS", "PN")
smpls <- smpls_df$Sample[is.element(smpls_df$HGCC, HGCCgroups)]
smpls <- sample(smpls, length(smpls), replace = FALSE)
```

We selected ```r length(smpls)``` samples for the analysis.
```{r metadata-4, eval = TRUE}
smpls
```

Retrieve metadata for selected samples. Create a column with `Case` and `Normal` conditions. 
```{r metadata-5, eval = TRUE}
smpls_df <- smpls_df[match(smpls, smpls_df$Sample),]
smpls_df$CaseControl <- smpls_df$HGCC
smpls_df$CaseControl[smpls_df$CaseControl != "Normal"] <- "Case"
```

```{r metadata-6, eval = TRUE}
write.table(x = smpls_df, file = "Glioblastoma_metadata.txt", sep = "\t", 
            quote = FALSE, row.names = FALSE)
```

Make factors of the `CaseControl` variables. It will be use for downstream analyses.
```{r }
sample_group <- as.factor(smpls_df$CaseControl)
```

## Import Glioblastoma mRNA data

Import mRNA data and make a raw counts matrix.
```{r mRNA-1, eval = TRUE}
expdat_mRNA <- lapply(X = smpls, FUN = mRNA_raw_import, mRNA_dir)
expdat_mRNA <- do.call(rbind, expdat_mRNA)
expdat_mRNA <- reshape2::dcast(expdat_mRNA, formula = gene_id ~ sample)
rownames(expdat_mRNA) <- expdat_mRNA$gene_id
expdat_mRNA <- as.matrix(expdat_mRNA[, -1])
expdat_mRNA[c(1:5), c(1:7)]
```

Tidy up feature names and order sample columns.
```{r mRNA-2, eval = TRUE}
expdat_mRNA <- mRNA_stripVersion(expdat_mRNA)
expdat_mRNA <- expdat_mRNA[,smpls]
expdat_mRNA[c(1:5), c(1:7)]
```

The raw counts matrix is saved into a .csv file to facilitate the loading.
```{r mRNA-3, eval = TRUE}
write.table(x = expdat_mRNA, file = file.path(mRNA_dir, "mRNA.csv"), 
            sep = ",", row.names = TRUE, col.names = TRUE)
```

The mRNA dataset contains `r nrow(expdat_mRNA)` features (genes) and `r ncol(expdat_mRNA)` samples.

## Import Glioblastoma DNAme data

Read DNAme data file. You will get warnings because the file contains some corrupted lines. We will remove these lines after.
```{r DNAme-1, eval = TRUE, warning = FALSE}
expdat_DNAme <- strsplit(
    readLines(file.path(DNAme_dir, "5mC_bVals_HGCC.txt"), n = -1),
    split = "\t"
)
```

Reshape the DNAme data object and keep only rows with the correct number of elements.
```{r DNAme-2, eval = TRUE}
DNAme_smpls <- expdat_DNAme[[1]]
correct_length <- 15
expdat_DNAme_chk <- lapply(expdat_DNAme, function(x) length(x) == correct_length)
expdat_DNAme <- expdat_DNAme[unlist(expdat_DNAme_chk)]
expdat_DNAme <- do.call(rbind, expdat_DNAme)
```

Make sure only rows with probe IDs are kept
```{r DNAme-3, eval = TRUE}
expdat_DNAme <- 
    expdat_DNAme[is.element(substr(expdat_DNAme[,1], 1, 2), c("cg", "ch")),]
```

Move sample IDs to the row name columns and add sample names.
```{r DNAme-4, eval = TRUE}
rownames(expdat_DNAme) <- expdat_DNAme[,1]
expdat_DNAme <- expdat_DNAme[,-1]
colnames(expdat_DNAme) <- DNAme_smpls
expdat_DNAme[c(1:5), c(1:3)]
```

Then, values need to be convert into numeric values. Don't worry, the introduced NA are in the `e12` column and we will remove it.
```{r DNAme-5, eval = TRUE, warning = FALSE, message = FALSE}
expdat_DNAme <- matrix(as.numeric(expdat_DNAme), 
                       ncol = ncol(expdat_DNAme),
                       dimnames = list(rownames(expdat_DNAme),
                                       colnames(expdat_DNAme)))
```

To analysis the DNAme data, values need to be converted into M-Values.
```{r DNAme-6, eval = TRUE}
expdat_DNAme <- sesame::BetaValueToMValue(expdat_DNAme)
```

Finally, we remove samples that are not in the `smpls` variable and order sample columns. 
```{r DNAme-7, eval = TRUE}
expdat_DNAme <- expdat_DNAme[,smpls]
```

The M-Values matrix is saved into a .csv file to facilitate the loading.
```{r DNAme-8, eval = TRUE}
write.table(x = expdat_DNAme, file = file.path(DNAme_dir, "DNAme.csv"), 
            sep = ",", row.names = TRUE, col.names = TRUE)
```

The DNAme dataset contains `r nrow(expdat_DNAme)` features (probes) and `r ncol(expdat_DNAme)` samples.


```{r}
smpls_raw <- sub("-", ".", smpls)
```

# Learning dataset

Download the learning dataset factorization files from [zenodo](https://zenodo.org/records/10848217):

- TCGA dataset factorization: [Lrn_5000D_Fctrzn_100k_001TH.zip](https://zenodo.org/records/10848217/files/Lrn_5000D_Fctrzn_100K_001TH.zip?download=1)
- Metadata file from TCGA data processing: [expdat_meta.rds](https://zenodo.org/records/10848217/files/expdat_meta.rds?download=1)

Here, factorization data are stored in the `Lrn_5000D_Fctrzn_100K_001TH/` folder and metadata in the `Lrn_5000D/` folder.
```{r Lrn_files, eval = TRUE}
Lrn_dir <- "Lrn_5000D/"
LrnFctrn_dir <- "Lrn_5000D_Fctrzn_100K_001TH/"
```

## Import learning dataset metadata

Import metadata of the learning dataset creation. 
```{r Lrn-1, eval = TRUE}
expdat_meta_Lrn <- readRDS(file.path(Lrn_dir, "expdat_meta_Lrn.rds"))
```

Import the learning dataset factorization model, created using `MOFA2`.
```{r Lrn-2, eval = TRUE}
InputModel <- file.path(LrnFctrn_dir, "Model.hdf5")
Fctrzn <- load_model(file = InputModel)
```

For the preprocessing step of the target dataset, you have two choices:

- `CenterTrg <- TRUE`: center the target dataset values
- `CenterTrg <- FALSE`: leave the values uncentered and use instead the estimated learning dataset intercepts

```{r Lrn-3, eval = TRUE}
CenterTrg <- FALSE
```

Initialize values from learning dataset factorization:

- `viewsLrn`: names of the learning dataset view (e.g. `r Fctrzn@data_options$views`)
- `likelihoodsLrn`: the defined likelihood of each view (e.g. `r Fctrzn@model_options$likelihoods`)
- `MLrn`: number of views in the learning dataset (`r Fctrzn@dimensions$M`)

```{r Lrn-4, eval = TRUE}
viewsLrn <- Fctrzn@data_options$views
likelihoodsLrn <- Fctrzn@model_options$likelihoods
MLrn <- Fctrzn@dimensions$M
```

In this use case, to initialize or caluclate the expected values, we use values from the leaning dataset factorization model generated using MOFA2.

- Initialize the `Tau` values using the one from the learning dataset model
- The log(Tau) values `TauLn` are calculated using the `Tau` values
- The squared weight values `WSq` are calculated using the weight values `W` 
- The weight intercept values `W0` are loaded from the estimated intercept file (`CenterTrg <- FALSE`) 

```{r Lrn-5, eval = TRUE}
Fctrzn@expectations[["Tau"]] <- Tau_init(viewsLrn, Fctrzn, InputModel)
Fctrzn@expectations[["TauLn"]] <- sapply(viewsLrn, TauLn_calculation, likelihoodsLrn, Fctrzn)
Fctrzn@expectations[["WSq"]] <- sapply(viewsLrn, WSq_calculation, Fctrzn)
Fctrzn@expectations[["W0"]] <- sapply(viewsLrn, W0_calculation, CenterTrg, Fctrzn, LrnFctrn_dir)
```


# Transfer learning using MOTL
# Downstream analysis



```{r }
library("MOTL")

library("fgsea")
library("RColorBrewer")
library("ggplot2")
library("pheatmap")
library("msigdbr")

```

# Target dataset

data are saved into `Glioblastoma/`
```{r wd}
glio_dir <- "Glioblastoma/"
Lrn_dir <- "Lrn_5000D/"
LrnFctrn_dir <- "Lrn_5000D_Fctrzn_100K_001TH/"
```


# Prepare datasets

If `CenterTrg = FALSE`, estimated intercepts are used for the normalization.
```{r parameters}
CenterTrg <- FALSE
```



## Target dataset: Glioblastoma data

```{r}
expdat_mRNA <- mRNA_addVersion(expdat = expdat_mRNA, Lrndat = Fctrzn@expectations$W$mRNA)
```

create a list of all target dataset omics data
```{r}
YTrg_list <- list(mRNA = expdat_mRNA, DNAme = expdat_DNAme)
```

harmonize the vectors of views and likelihoods
```{r}
viewsTrg <- names(YTrg_list)
views <- viewsLrn[is.element(viewsLrn, viewsTrg)]
likelihoods <- likelihoodsLrn[views]
```

Prepare YTrg data
```{r}
YTrg_list <- TargetDataPreparation(views = views, YTrg_list = YTrg_list, 
                                  Fctrzn = Fctrzn, smpls = smpls, expdat_meta_Lrn = expdat_meta_Lrn,
                                  normalization = 'LrnGeoMeans', transformation = TRUE)
```

Init parameters
```{r}
TL_param <- initTransferLearningParamaters(YTrg = YTrg_list, views = views, 
                                          Fctrzn = Fctrzn, likelihoods = likelihoods
                                          )
```

## Transfer learning

defining output folder here - this should be an input to the funciton
```{r}
TL_outDir <- file.path(glio_dir, "MOTL_results")
if(!dir.exists(TL_outDir)){
  dir.create(TL_outDir, showWarnings = FALSE, recursive = TRUE)
}
```

Parameters
```{r}
minFactors <- 13 ## floor when dropping factors - number of samples in evaluations
StartDropFactor <- 1 # after which iteration to start dropping factors
FreqDropFactor <- 1 ## how often to drop factors
StartELBO <- 1 #which iteration to start checking ELBO on, excl initiation iteration
FreqELBO <- 5 #how often to assess the ELBO
DropFactorTH <- 0.01 # factor with lowest max variance, that is less than this, is dropped
MaxIterations <- 10000
MinIterations <- 2 # as per MOFA with defaults in python, >= 2 (hardcoded floor), exl initial setup
ConvergenceIts <- 2 # as per MOFA python defaults
ConvergenceTH <- 0.0005 # default for MOFA - fast option
# PoisRateCstnt <- 0.0001 ## amount to add to the poison rate function to avoid errors
```

Run MOTL
```{r, message = FALSE}
TL_data <- transferLearning_function(TL_param = TL_param, views = views, Fctrzn = Fctrzn,
                                    likelihoods = likelihoods,
                                    MaxIterations = MaxIterations, 
                                    MinIterations = MinIterations, minFactors = minFactors, 
                                    StartDropFactor = StartDropFactor, FreqDropFactor = FreqDropFactor, 
                                    StartELBO = StartELBO, FreqELBO = FreqELBO, DropFactorTH = DropFactorTH, 
                                    ConvergenceIts = ConvergenceIts, ConvergenceTH = ConvergenceTH, 
                                    CenterTrg = CenterTrg, outputDir = TL_outDir)
```

Save the metadata
```{r}
TL_meta_data <- list(
  "Seed" = Seed,
  "CenterTrg" = CenterTrg,
  "StartDropFactor" = StartDropFactor,
  "FreqDropFactor" = FreqDropFactor,
  "DropFactorTH" = DropFactorTH,
  "minFactors" = minFactors,
  "MinIterations" = MinIterations,
  "MaxIterations" = MaxIterations,
  "StartELBO" = StartELBO,
  "FreqELBO" = FreqELBO,
  "ConvergenceTH" = ConvergenceTH,
  "ConvergenceIts" = ConvergenceIts
)
saveRDS(TL_meta_data, file.path(TL_outDir, "TL_VI_meta_data.rds"))
```

## Downstream analysis

Find associated pathways and processes

```{r}
TL_data <- readRDS(file.path(TL_outDir, "TL_data.rds"))
```

```{r}
YGauss <- TL_data$YGauss
ZMu_0 <- TL_data$ZMu_0
ZMu <- TL_data$ZMu
Fctrzn_Lrn_W0 <- TL_data$Fctrzn_Lrn_W0
Fctrzn_Lrn_W <- TL_data$Fctrzn_Lrn_W
views <- names(YGauss)
factorNames <- colnames(ZMu)
VarExpl <- TL_data$VarExpl
```

Test for differentially active factors

```{r}
minSamples <- min(unlist(table(sample_group)))

```

```{r}
DifFactTester <- function(x, minSamples, sample_group) {
  if (minSamples >= 4) {
    wilcox.test(x ~ sample_group)$p.value
  } else {
    t.test(x ~ sample_group)$p.value
  }
}
```

```{r}
Z_pv <- apply(X = ZMu, MARGIN = 2, FUN = DifFactTester, minSamples = minSamples, sample_group = sample_group)
Z_pv <- p.adjust(Z_pv, method = "BH")
Z_pv_sig <- which(Z_pv < 0.05)
names(Z_pv_sig) <- names(Z_pv)[Z_pv_sig]
```

retrieve variance explained for mRNA
```{r}
Z_mRNA_rel <- which(VarExpl[,"mRNA"] > 0.01)
names(Z_mRNA_rel) <- rownames(VarExpl)[Z_mRNA_rel]
```

```{r}
## GSEA mRNA weights for factors that are differentially active and relevant to mRNA
####

#  scale and tidy row names of mRNA W matrix
GroupsTested <- paste(levels(sample_group), collapse = "_vs_")
ScaleW <- function(x) {
  x/norm(matrix(x), type = "F")
}
W_mRNA <- t(apply(Fctrzn_Lrn_W[["mRNA"]], 1, ScaleW))
W_mRNA <- mRNA_stripVersion(W_mRNA)
```

```{r}
# factors and genestes to test
kToTest <- names(Z_pv_sig)[is.element(names(Z_pv_sig), names(Z_mRNA_rel))]
GenesetCatToTest <- c("C2", "C2", "C5", "C5")
GenesetSubCatToTest <- c("CP:KEGG", "CP:REACTOME", "GO:BP", "GO:CC")

ToTest <- data.frame(
  k = rep(kToTest, each = length(GenesetCatToTest)),
  category = rep(GenesetCatToTest, times = length(kToTest)),
  subcategory = rep(GenesetSubCatToTest, times = length(kToTest))
)


SigGeneSets <- apply(X = ToTest, MARGIN = 1, FUN = function(ToTestVec, WMat = W_mRNA, Groups = GroupsTested){
  
  k <- as.character(ToTestVec[1])
  category <- as.character(ToTestVec[2])
  subcategory <- as.character(ToTestVec[3])
  rankings <- WMat[,k]
  names(rankings) <- rownames(WMat)
  
  gene_sets <- msigdbr(
    species = "Homo sapiens", 
    category = category, 
    subcategory = subcategory
  )
  
  gene_sets <- gene_sets %>% 
    dplyr::distinct(gs_name, ensembl_gene) %>% 
    as.data.frame()
  
  gene_sets <- split(
    x = gene_sets$ensembl_gene, 
    f = gene_sets$gs_name
  )
  # set seed
  Seed <- 1234567
  mode(Seed) <- "integer"
  set.seed(Seed)
  
  # run fgsea
  fgseaRes <- fgsea(
    pathways = gene_sets,
    stats = rankings,
    minSize = 15,
    maxSize = 500
  )
  
  fgseaRes <- fgseaRes %>% 
    dplyr::select(pathway, padj, ES, NES, size) %>% 
    as.data.frame()
  
  fgseaRes <- fgseaRes[fgseaRes$padj < 0.01 & !is.na(fgseaRes$padj),]
  fgseaRes$k <- rep(k, nrow(fgseaRes))
  fgseaRes$subcategory <- rep(subcategory, nrow(fgseaRes))
  fgseaRes$Groups <- rep(Groups, nrow(fgseaRes))
  
  return(fgseaRes)
  
})

SigGeneSets <- do.call(rbind, SigGeneSets)

## export as csv
write.table(SigGeneSets,
            file.path(TL_OutDir, paste0('SigGeneSets','_',GroupsTested,'.csv')),
            quote = FALSE, sep=',', na='', row.names = FALSE, col.names = TRUE)
```







```{r}
#order the samples
smpls_df <- smpls_df[order(smpls_df$CaseControl, smpls_df$HGCC, smpls_df$Sample),]
smpls_df$CaseControl[smpls_df$CaseControl == "Case"] <-"pd-GBSC"
```

```{r}
# motl factorization
TL_fctrzn <- readRDS(file.path(TL_outDir, "TL_data.rds"))
```

```{r}
Zmu_MOTL <- t(TL_fctrzn$ZMu[smpls_df$Sample,])

heatMapAnno <- data.frame(
  SubType = as.factor(smpls_df$HGCC),
  Type = as.factor(smpls_df$CaseControl)
)

heatMapAnno$SubType <- relevel(heatMapAnno$SubType, ref = "Normal")
heatMapAnno$Type <- relevel(heatMapAnno$Type, ref = "Normal")
rownames(heatMapAnno) <- smpls_df$Sample

annot_colors <- list(SubType = brewer.pal(5, "Set2")[c(1,3,4,5)],
                   Type = brewer.pal(5, "Set2")[1:2])

names(annot_colors$SubType) <- levels(heatMapAnno$SubType)
names(annot_colors$Type) <- levels(heatMapAnno$Type)

pheatmap(Zmu_MOTL, 
         scale = "row", 
         border_color = NA,
                  # cellwidth = 9,
                  # cellheight = 3,
                  # fontsize = 6,
                  angle_col = 45,
                  cluster_rows = TRUE,
                  cluster_cols = TRUE,
                  # clustering_method = "single",
                  treeheight_row = 0,
                  treeheight_col = 10,
                  cutree_cols = 1,
                  legend = FALSE,
                  show_rownames = FALSE,
                  show_colnames = TRUE,
                  annotation_col = heatMapAnno,
                  annotation_colors = annot_colors,
                  annotation_names_col = TRUE,
                  annotation_legend = TRUE)
```



