---
title: "UseCase_Glioblastoma"
output: 
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{UseCase_Glioblastoma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MOTL)
Seed = 1234567
mode(Seed) = 'integer'
set.seed(Seed)
```

Helper functions to import

# Target dataset

data are saved into `../data-raw/Glioblastoma`
```{r wd}
glio_dir <- "data-raw/Glioblastoma"
Lrn_dir <- "data-raw/Lrn_5000D"
LrnFctrn_dir <- "data-raw/Lrn_5000D_Fctrzn_100K_001TH"
```

## Import mRNA data

Metadata of mRNA dataset. Replace `.` by `-`.
```{r}
smpls_df <- read.table(file.path(glio_dir, "02_RNAseq/phenodata_RNAseq.tsv"), header = TRUE)
smpls_df$Sample <- sub("[.]", "-", smpls_df$Sample)
smpls_df$Sample <- sub("X", "", smpls_df$Sample)
```

Only interested in 3 types of cancer. Randomly order sample names.
```{r}
HGCCgroups <- c("Normal", "CL", "MS", "PN")
smpls <- smpls_df$Sample[is.element(smpls_df$HGCC, HGCCgroups)]
smpls <- sample(smpls, length(smpls), replace = FALSE)
```

Retrieve metadata for selected samples. Create a column with `Case` and `Normal` conditions. 
```{r}
smpls_df <- smpls_df[match(smpls, smpls_df$Sample),]
smpls_df$CaseControl <- smpls_df$HGCC
smpls_df$CaseControl[smpls_df$CaseControl!="Normal"] <- "Case"
```

Make factors of the `CaseControl` variables. It will be use for downstream analyses.
```{r}
sample_group <- as.factor(smpls_df$CaseControl)
```

Import mRNA data and make a raw counts matrix.

```{r}
smpls_raw <- sub("-", ".", smpls)
```


```{r}
mRNA_raw_import <- function(smpls, InRawCountsDir){
  # import raw counts line by line for each sample
  mRNA_trg_tmp <- strsplit(
    readLines(file.path(InRawCountsDir, paste0(smpls,'.genes.results')), n = -1), 
    split = "\t"
  )
  
  correct_length <- 7
  mRNA_trg_tmp_chk <- lapply(mRNA_trg_tmp, function(x) length(x) == correct_length)
  mRNA_trg_tmp <- mRNA_trg_tmp[unlist(mRNA_trg_tmp_chk)]
  mRNA_trg_tmp <- do.call(rbind, mRNA_trg_tmp)
  colnames(mRNA_trg_tmp) <- mRNA_trg_tmp[1,]
  mRNA_trg_tmp <- mRNA_trg_tmp[-1,]
  
  mRNA_trg_tmp <- data.frame(
    gene_id = mRNA_trg_tmp[,'gene_id'],
    sample = smpls,
    expected_count = round(as.numeric(mRNA_trg_tmp[,'expected_count']))
  )
  
  return(mRNA_trg_tmp)
}
```



```{r}
InRawCountsDir <- file.path(glio_dir, "02_RNAseq/CountData/")
expdat_mRNA <- lapply(X = smpls, FUN = mRNA_raw_import, InRawCountsDir)
expdat_mRNA <- do.call(rbind, expdat_mRNA)
expdat_mRNA <- reshape(expdat_mRNA,
                      idvar = "gene_id",
                      timevar = "sample",
                      direction = "wide")
rownames(expdat_mRNA) <- expdat_mRNA$gene_id
expdat_mRNA <- as.matrix(expdat_mRNA[, -1])
colnames(expdat_mRNA) <- substr(colnames(expdat_mRNA), 16, nchar(colnames(expdat_mRNA)))
```


```{r}
mRNA_stripVersion <- function(expdat){
  # strip version and gene symbol from ensembl ids
  vStart <- regexpr('[.]', rownames(expdat))
  suffixStart <- regexpr('_PAR_Y', rownames(expdat))
  suffixEnd <- suffixStart -1 + attr(regexpr('_PAR_Y', rownames(expdat)), "match.length")
  rownames(expdat) <- paste0(
    substr(rownames(expdat), 1, vStart-1),
    substr(rownames(expdat), suffixStart, suffixEnd)
  )
  # return tidied up matrix
  return(expdat)
}
```

Tidy up features and sample names
```{r}
expdat_mRNA <- mRNA_stripVersion(expdat_mRNA)
expdat_mRNA <- expdat_mRNA[,smpls]
```

```{r}
write.table(x = expdat_mRNA, file = file.path(glio_dir, "02_RNAseq/mRNA.csv"), sep = ",", row.names = TRUE, col.names = TRUE)
```

## DNAme

Read file
```{r}
expdat_DNAme <- strsplit(
  readLines(file.path(glio_dir, "03_DNAmethylation/5mC_bVals_HGCC.txt"), n = -1), 
  split = "\t"
)
```

Keep sample names and tidy
```{r}
DNAme_smpls <- expdat_DNAme[[1]]
```

Only keep lines with spefified number of elements
```{r}
correct_length <- 15
expdat_DNAme_chk <- lapply(expdat_DNAme, function(x) length(x) == correct_length)
expdat_DNAme <- expdat_DNAme[unlist(expdat_DNAme_chk)]
expdat_DNAme <- do.call(rbind, expdat_DNAme)
```

make sure only rows with probe ids are kept
```{r}
expdat_DNAme <- expdat_DNAme[is.element(substr(expdat_DNAme[,1], 1, 2), c('cg', 'ch')),]
```

move ids to row names
```{r}
rownames(expdat_DNAme) <- expdat_DNAme[,1]
expdat_DNAme <- expdat_DNAme[,-1]
colnames(expdat_DNAme) <- DNAme_smpls
```

convert to numeric
```{r}
expdat_DNAme = matrix(
  as.numeric(expdat_DNAme),
  ncol = ncol(expdat_DNAme),
  dimnames = list(
    rownames(expdat_DNAme),
    colnames(expdat_DNAme)
  )
)
```

convert to M-Values
```{r}
expdat_DNAme = sesame::BetaValueToMValue(expdat_DNAme)
```

filter and order samples
```{r}
expdat_DNAme = expdat_DNAme[,smpls]
```

```{r}
write.table(x = expdat_DNAme, file = file.path(glio_dir, "03_DNAmethylation/DNAme.csv"), sep = ",", row.names = TRUE, col.names = TRUE)
```


# Prepare datasets

If `CenterTrg = FALSE`, estimated intercepts are used for the normalization.
```{r parameters}
CenterTrg <- FALSE
```

## Learning dataset

Import metadata
```{r Lrn-1}
expdat_meta_Lrn <- readRDS(file.path(Lrn_dir, "expdat_meta_Lrn.rds"))

```

```{r}
InputModel <- file.path(LrnFctrn_dir, "Model.hdf5")
Fctrzn <- load_model(file = InputModel)
```

```{r}
viewsLrn <- Fctrzn@data_options$views
likelihoodsLrn <- Fctrzn@model_options$likelihoods
MLrn <- Fctrzn@dimensions$M
```

LOAD IN THE EXPECTATIONS FOR TAU AND REORDER FOR CONSISTENCY
```{r}
Fctrzn@expectations[["Tau"]] <- Tau_init(viewsLrn, Fctrzn, InputModel)
```

E[log(tau)] CALCULATION (ONLY RELEVANT FOR GAUSSIAN DATA)
```{r}
Fctrzn@expectations[["TauLn"]] <- sapply(viewsLrn, TauLn_calculation, likelihoodsLrn, Fctrzn)
```

load or calculate E[W^2] values
```{r}
Fctrzn@expectations[["WSq"]] <- sapply(viewsLrn, WSq_calculation, Fctrzn)
```

import the W intercepts if not centering, create vectors containing 0s if centering
```{r}
Fctrzn@expectations[["W0"]] <- sapply(viewsLrn, W0_calculation, CenterTrg, Fctrzn, LrnFctrn_dir)
```

## Target dataset: Glioblastoma data

```{r}
expdat_mRNA <- mRNA_addVersion(expdat = expdat_mRNA, Lrndat = Fctrzn@expectations$W$mRNA)
```

create a list of all target dataset omics data
```{r}
YTrg_list <- list(mRNA = expdat_mRNA, DNAme = expdat_DNAme)
```

harmonize the vectors of views and likelihoods
```{r}
viewsTrg <- names(YTrg_list)
views <- viewsLrn[is.element(viewsLrn, viewsTrg)]
likelihoods <- likelihoodsLrn[views]
```

Prepare YTrg data
```{r}
YTrg_list <- TargetDataPreparation(views = views, YTrg_list = YTrg_list, 
                                  Fctrzn = Fctrzn, smpls = smpls, expdat_meta_Lrn = expdat_meta_Lrn,
                                  normalization = 'LrnGeoMeans', transformation = TRUE)
```

Init parameters
```{r}
TL_param <- initTransferLearningParamaters(YTrg = YTrg_list, views = views, 
                                          Fctrzn = Fctrzn, likelihoods = likelihoods
                                          )
```

## Transfer learning

defining output folder here - this should be an input to the funciton
```{r}
TL_outDir <- file.path(glio_dir, "MOTL_results")
if(!dir.exists(TL_outDir)){
  dir.create(TL_outDir, showWarnings = FALSE, recursive = TRUE)
}
```

Parameters
```{r}
minFactors <- 13 ## floor when dropping factors - number of samples in evaluations
StartDropFactor <- 1 # after which iteration to start dropping factors
FreqDropFactor <- 1 ## how often to drop factors
StartELBO <- 1 #which iteration to start checking ELBO on, excl initiation iteration
FreqELBO <- 5 #how often to assess the ELBO
DropFactorTH <- 0.01 # factor with lowest max variance, that is less than this, is dropped
MaxIterations <- 10000
MinIterations <- 2 # as per MOFA with defaults in python, >= 2 (hardcoded floor), exl initial setup
ConvergenceIts <- 2 # as per MOFA python defaults
ConvergenceTH <- 0.0005 # default for MOFA - fast option
# PoisRateCstnt <- 0.0001 ## amount to add to the poison rate function to avoid errors
```

Run MOTL
```{r, message = FALSE}
TL_data <- transferLearning_function(TL_param = TL_param, views = views, Fctrzn = Fctrzn,
                                    likelihoods = likelihoods,
                                    MaxIterations = MaxIterations, 
                                    MinIterations = MinIterations, minFactors = minFactors, 
                                    StartDropFactor = StartDropFactor, FreqDropFactor = FreqDropFactor, 
                                    StartELBO = StartELBO, FreqELBO = FreqELBO, DropFactorTH = DropFactorTH, 
                                    ConvergenceIts = ConvergenceIts, ConvergenceTH = ConvergenceTH, 
                                    CenterTrg = CenterTrg, outputDir = TL_outDir)
```

Save the metadata
```{r}
TL_meta_data <- list(
  "Seed" = Seed,
  "CenterTrg" = CenterTrg,
  "StartDropFactor" = StartDropFactor,
  "FreqDropFactor" = FreqDropFactor,
  "DropFactorTH" = DropFactorTH,
  "minFactors" = minFactors,
  "MinIterations" = MinIterations,
  "MaxIterations" = MaxIterations,
  "StartELBO" = StartELBO,
  "FreqELBO" = FreqELBO,
  "ConvergenceTH" = ConvergenceTH,
  "ConvergenceIts" = ConvergenceIts
)
saveRDS(TL_meta_data, file.path(TL_outDir, "TL_VI_meta_data.rds"))
```

## Downstream analysis
